\documentclass{article}
\usepackage{amsmath}
\usepackage{verbatim}
%\usepackage{listings}
\allowdisplaybreaks

\begin{document}
\title{Coulomb lattice summation}
\author{}
\maketitle

We use Coulomb lattice summation to calculate the matrix elements of the
modified potential $\tilde{V}$
\begin{align}
  \tilde{V}_{\mu\nu}^{0J}
  &=\langle \mu^0|\tilde{V}|\nu^J\rangle
  =\int \mu^0(\mathbf{r}) \tilde{V}(\mathbf{r}) \nu^J(\mathbf{r}) d\mathbf{r}^3
  \\
  \tilde{V}(\mathbf{r})
  &=\int \frac{\tilde{\rho}(\mathbf{r}')}{|\mathbf{r}-\mathbf{r}'|}
  d{\mathbf{r}'}^3
  \label{}
\end{align}
where the superscripts $0$, $J$ of $\mu^0$ and $\nu^J$ labels the indices of
cells, $\nu^J$ means the $\nu$-th AO of $J$-th cell.
Therefore the matrix representation of $V$ is an array with 
$n_\text{AO} \times n_\text{AO} \times n_\text{UnitCells}$ elements.
$\tilde{V}$ is based on the screened electron density $\tilde{\rho}(\mathbf{r})$
which is the difference between the electron density $\rho(\mathbf{r})$ and
the averaged density $\rho_0$
\begin{align}
  &\tilde{\rho}(\mathbf{r})
  =\rho(\mathbf{r}) - \rho_0
  \\
  &\int_{\mathbf{r}\in C} \rho(\mathbf{r}) d\mathbf{r}^3 = N
  \\
  &\rho_0 = \frac{N}{\Omega},\quad
  \int_{\mathbf{r}\in C} \tilde{\rho}(\mathbf{r}) d\mathbf{r}^3 = 0
  \label{}
\end{align}
where the integral region $C$ is constrained to a unit cell,
$\Omega$ is the volume of a unit cell.

\section{Formulas}
Basically, the algorithm contains the following steps:
\begin{itemize}
  \item Calculate the screened density $\tilde{\rho}(\mathbf{r})$ at each mesh
    grid of real space;
  \item Use FFT to transform $\tilde{\rho}(\mathbf{r})$ to its representation
    in $\mathbf{k}$-space $\tilde{\rho}(\mathbf{k})$;
  \item Solve the Poisson equation in $\mathbf{k}$-space and get the
    $\mathbf{k}$-space potential $\tilde{V}(\mathbf{k})$;
  \item Back transform $\tilde{V}(\mathbf{k})$ to the real space
    representation $\tilde{V}(\mathbf{r})$;
  \item Calculate the matrix elements $V_{\mu\nu}^{0J}$ numerically in terms
    of the real space mesh grids.
\end{itemize}

The screened density in the real space reads
\begin{align}
  \tilde{\rho}(\mathbf{r}) = \sum_{\mu \nu IJ}D_{\mu\nu}^{IJ}
  \mu^{I}(\mathbf{r})\nu^{J}(\mathbf{r}) - \rho_0
  \label{eq:rho:r}
\end{align}
The superscripts $I$ and $J$ indicates the indices of cells.
The density matrix $D$ has the translation symmetry
\begin{equation}
  D_{\mu\nu}^{IJ} = D_{\mu\nu}^{0,J-I}
  \label{}
\end{equation}
where the superscripts $J-I$ means the cell which corresponds to the lattice
vector pointing from $I$-th cell to $J$-th cell
\begin{equation}
  \mathbf{R}^{J-I} = \mathbf{R}^J - \mathbf{R}^I
  \label{}
\end{equation}
The periodic feature of basis function allows us to transform the basis of $I$-th cell to
the basis of $0$-th cell
\begin{equation}
  \mu^I(\mathbf{r}) = \mu^0(\mathbf{r}-\mathbf{R}^I)
  \label{}
\end{equation}
After considering the periodic condition and the translation symmetry,
we can rewrite the expression of density Eq. \eqref{eq:rho:r} as
\begin{align}
  \tilde{\rho}(\mathbf{r})
  &=\sum_{\mu\nu IJ}D_{\mu\nu}^{I,I+J}
  \mu^I(\mathbf{r})\nu^{I+J}(\mathbf{r}) - \rho_0
  =\sum_{I} \rho^I(\mathbf{r}) - \rho_0
  \label{eq:rho:div} \\
  \rho^I(\mathbf{r})
  &=\sum_{\mu\nu J}D_{\mu\nu}^{0J}
  \mu^I(\mathbf{r})\nu^{I+J}(\mathbf{r}) \notag\\
  &=\sum_{\mu\nu J}D_{\mu\nu}^{0J}
  \mu^0(\mathbf{r}-\mathbf{R}^I)\nu^0(\mathbf{r}-\mathbf{R}^{I+J})
  \label{eq:rho:ra}
\end{align}
$D_{\mu\nu}^{0J}$ yields a density matrix with
$n_\text{AO} \times n_\text{AO} \times n_\text{UnitCells}$ elements.
Eq. \eqref{eq:rho:div} decomposes the total density into the individual
cellular contributions.
Every single cell $\rho^I(\mathbf{r})$ contributes to the density
distribution over the whole space.
Due to the periodic condition, 
we have the same density distribution for each single cell.
So their contributions to the total density can be refined to the contribution
of $0$-th cell
\begin{align}
  \tilde{\rho}(\mathbf{r})
  &=\sum_{I}\rho^I(\mathbf{r}) - \rho_0
  = \sum_{I}\rho^0(\mathbf{r}-\mathbf{R}^I) - \rho_0
  \label{eq:rho:r0} \\
  \rho^0(\mathbf{r})
  &=\sum_{\mu\nu J}D_{\mu\nu}^{0J}
  \mu^0(\mathbf{r})\nu^0(\mathbf{r}-\mathbf{R}^{J})
  = \sum_{\mu\nu J}D_{\mu\nu}^{0J}
  \mu^0(\mathbf{r})\nu^J(\mathbf{r})
\end{align}

The FFT of $\tilde{\rho}$ in the whole real space reads
\begin{align}
  \tilde{\rho}(\mathbf{k})
  =\int d\mathbf{r}^3 \tilde{\rho}(\mathbf{r}) e^{-i\mathbf{k}\cdot\mathbf{r}}
  =\sum_{\mathbf{r}} \tilde{\rho}(\mathbf{r})
  e^{-i\mathbf{k}\cdot \mathbf{r}}
  =\sum_{\mathbf{r} I}(\rho^I(\mathbf{r}) - \rho^0)
  e^{-i\mathbf{k}\cdot \mathbf{r}}
  \label{}
\end{align}
The subscription $\mathbf{r}$ of $\sum$ is the index of grid in the real
space.  It indicates that the integral is calculated numerically in
terms of the mesh grids.
Due to the periodic condition,
if the density distribution in unit cell was evaluated,
we can know the values in all the other cells.
Therefore, we can confine the FFT in whole space to the FFT in a finite space,
say the volume of one cell,
\begin{align}
  \tilde{\rho}(\mathbf{k})
  =\int_{\mathbf{r}\in C} d\mathbf{r}^3
  \tilde{\rho}(\mathbf{r}) e^{-i\mathbf{k}\cdot\mathbf{r}}
  =\sum_{\mathbf{r}\in C}
  \tilde{\rho}(\mathbf{r}) e^{-i\mathbf{k}\cdot \mathbf{r}}
  \label{eq:rhok:unitcell}
\end{align}

The Poisson equation in real space reads
\begin{equation}
  \nabla^2 \tilde{V}(\mathbf{r}) = 4\pi \tilde{\rho}(\mathbf{r})
  \label{eq:poisson}
\end{equation}
Solving the Poisson equation in $\mathbf{k}$-space leads to
the $\mathbf{k}$-space representation of the potential
\begin{equation}
  \tilde{V}(\mathbf{k})=\frac{4\pi\tilde{\rho}(\mathbf{k})}{k^2}
  \label{}
\end{equation}
$\mathbf{k}$ vector can be represented as a vector in terms of the reciprocal
basis $\mathbf{b}$
\begin{align}
  \mathbf{b}_1 &= (b_{1x},b_{1y},b_{1z})\\
  \mathbf{b}_2 &= (b_{2x},b_{2y},b_{2z})\\
  \mathbf{b}_3 &= (b_{3x},b_{3y},b_{3z})\\
  \mathbf{k}
  &=\mathbf{b}_1 k_1 + \mathbf{b}_2 k_2 + \mathbf{b}_3 k_3
  = k_x \mathbf{e}_x + k_y \mathbf{e}_y + k_z \mathbf{e}_z
  = (k_x, k_y, k_z) \\
  k_x &= b_{1x}k_1 + b_{2x}k_2 + b_{3x}k_3 \\
  k_y &= b_{1y}k_1 + b_{2y}k_2 + b_{3y}k_3 \\
  k_z &= b_{1z}k_1 + b_{2z}k_2 + b_{3z}k_3
  \label{}
\end{align}
If the reciprocal basis is non-orthogonal,
the denominator of Eq. \eqref{eq:poisson} is calculated via
\begin{equation}
  k^2 = k_x^2 + k_y^2 + k_z^2
  \label{}
\end{equation}

Carrying out the inverse FFT, we obtain the real space solution of the Poisson
equation \eqref{eq:poisson}
\begin{align}
  \tilde{V}(\mathbf{r})
  =\frac{1}{\Omega}\int_{\mathbf{k}\in C} d\mathbf{k}^3
  \tilde{V}(\mathbf{k}) e^{i\mathbf{k}\cdot\mathbf{r}}
  =\frac{1}{n_xn_yn_z}\sum_{\mathbf{k}\in C}
  \Phi(\mathbf{k}) e^{i\mathbf{k}\cdot \mathbf{r}}
  \label{eq:v:real}
\end{align}
where $n_x,n_y,n_z$ are the number of mesh grids along the Cartesian axes in
one cell.  We have already employed the 3D discrete Fourier transform (DFT)
and its inverse counterpart in the above equations
\begin{align}
  \tilde{\rho}(\mathbf{k}_x,\mathbf{k}_y,\mathbf{k}_z)
  &=\sum_{x=0}^{n_x-1} \sum_{y=0}^{n_y-1} \sum_{z=0}^{n_z-1}
  \tilde{\rho}(\mathbf{r}_x,\mathbf{r}_y,\mathbf{r}_z)
  \exp\left\{-2\pi i \mathbf{k}\cdot\mathbf{r}\right\},
  %\notag\\
  %&\quad \mathbf{k}_x \in [0,n_x-1],
  %\mathbf{k}_y=[0,n_y-1],\mathbf{k}_z \in 
  \\
  \tilde{\rho}(\mathbf{r}_x,\mathbf{r}_y,\mathbf{r}_z)
  &=\frac{1}{n_xn_yn_z}\sum_{x=0}^{n_x-1}
  \sum_{y=0}^{n_y-1} \sum_{z=0}^{n_z-1}
  \tilde{\rho}(\mathbf{k}_x,\mathbf{k}_y,\mathbf{k}_z)
  \exp\left\{2\pi i \mathbf{k}\cdot \mathbf{r}\right\}
  \label{}
\end{align}

Eq. \eqref{eq:v:real} gives the real space potential of one cell.
Using the translation symmetry, we can obtain the value of
$\tilde{V}(\mathbf{r})$ in the remaining cells
%\begin{equation}
%  \tilde{V}(\mathbf{r}) = \tilde{V}(\mathbf{r}-\mathbf{R}^I)
%  \label{}
%\end{equation}
\begin{equation}
  \tilde{V}(\mathbf{r}\in I) = \tilde{V}(\mathbf{r}-\mathbf{R}^I)
  \label{}
\end{equation}
Based on the periodic feature of the basis function and the potential, we
have
\begin{align}
  \tilde{V}_{\mu\nu}^{0J}
  &=\sum_{\mathbf{r}} \omega_\mathbf{r}
  \mu^0(\mathbf{r})\tilde{V}(\mathbf{r})\nu^J(\mathbf{r})
  \\
  &=\sum_{I}\sum_{\mathbf{r}\in I} \omega_\mathbf{r} \mu^0(\mathbf{r})
  \tilde{V}(\mathbf{r} - \mathbf{R}^I)\nu^0(\mathbf{r}-\mathbf{R}^J)
  \\
  &=\sum_{I}\sum_{\mathbf{r}\in 0} \omega_\mathbf{r}
  \mu^0(\mathbf{r}+\mathbf{R}^I)
  \tilde{V}(\mathbf{r})\nu^0(\mathbf{r} + \mathbf{R}^{I-J})
  =\sum_I \tilde{V}_{\mu\nu}^{0J[I]}
  \label{eq:v:r0} \\
  \tilde{V}_{\mu\nu}^{0J[I]}
  &=\sum_{\mathbf{r}\in 0} \omega_\mathbf{r} \mu^0(\mathbf{r}+\mathbf{R}^I)
  \tilde{V}(\mathbf{r})\nu^J(\mathbf{r} + \mathbf{R}^{I})
\end{align}
where $\omega_\mathbf{r}$ is the integration weight of grid.

For a unit cell, the averaged energy which corresponds to $\tilde{V}$ is
\begin{equation}
  \varepsilon = \int_{\mathbf{r}\in C} \rho(\mathbf{r})\tilde{V}(\mathbf{r}) d\mathbf{r}
  = \frac{1}{N}\sum_{\mu\nu J} D_{\mu\nu}^{0J} \tilde{V}_{\mu\nu}^{0J}
  = \sum_{\mu\nu} D_{\mu\nu}^{00} \tilde{V}_{\mu\nu}^{00}
  \label{}
\end{equation}
$\rho$ is the actual density, $N$ is the number of unit cells in a super cell.

In the above equations, the summation limits of AO indices $\mu,\nu,\dots$,
cell indices $I,J,\dots$, and mesh grid indices $\mathbf{r}$ which correspond
to the electron coordinates are not yet determined.
The indices $\mu$ and $\nu$ are always limited to the number of AOs in one cell.
In principle, the cell index $I$ should run over all unit cells in a
crystal.  The summation over $I$ can always be truncated.
In current implementation, we use a super cell to mimic the crystal,
i.e., we sum over all the unit cells' contribution in the super cell.
Since the 0-th cell locates at the super cell boundary, we rearrange the
super-cell structure to let the 0-th cell be in the center.
In Eq. \eqref{eq:rho:r0} and \eqref{eq:v:r0}, we constrain the electron
coordinate index $\mathbf{r}$ to be in the $0$-th cell.
The index of $\mathbf{r}$ thus go from 0 to the number of mesh grids in unit
cell which is
\begin{equation}
  n = n_x n_y n_z
  \label{}
\end{equation}


\section{Programming model}
We set up a class ?\verb$grid_3d$ to store the
grid informations. It includes the
\begin{itemize}
  \item the coordinates of mesh grids in one cell
  \item ??the value of basis on each grid
\end{itemize}

\subsection{API functions}
\begin{itemize}
\item
\begin{verbatim}
double cls::coul_matrix(const FSolidModel& solid,
        const FOpMatrix& den_mat, FOpMatrix& coul_mat);
\end{verbatim}
\begin{itemize}
  \item function to get the Coulomb matrix.
  \item input: \verb$den_mat$ in Fortran-contiguous order.
  \item output: \verb$coul_mat$ in Fortran-contiguous order.
  \item return: the energy in terms of the actual electron density $\rho$ and
    the modified potential $\tilde{V}$
    \begin{equation}
      \varepsilon = \int_{\mathbf{r}\in C} \rho(\mathbf{r})\tilde{V}(\mathbf{r}) d\mathbf{r}
      \label{}
    \end{equation}
\end{itemize}
%
\item 
\begin{verbatim}
void cls::set_num_grids_in_unit_cell(const int n_a1,
        const int n_a2, const int n_a3);
\end{verbatim}
\begin{itemize}
  \item uniform mesh grids are employed for numerical integration.
    Set the number of grids along the 3 lattice vectors in a unit cell.
    Default value is (20,20,20).
\end{itemize}
\end{itemize}

\section{Reference}
Mol. Phys. 77, 629

\end{document}
